---
title: "Spectra use case"
author: 
- name: "Johannes Rainer"
  affiliation: "Eurac Research, Bolzano, Italy; johannes.rainer@eurac.edu github: jorainer twitter: jo_rainer"
graphics: yes
output:
  BiocStyle::html_document:
    number_sections: true
    toc_float: true
    toc_depth: 2
---

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
library("BiocStyle")
library("knitr")
library("rmarkdown")
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7) 
```


# `Spectra`

-   Example use cases for the `Spectra` package and functionality.

## Import data from an *mzML* file

- What is a `Spectra` object?
- *Example*: Load an mzML from an LC-MS/MS (DDA) experiment.
```{r}
library(Spectra)
library(magrittr)
    
sps <- Spectra("data/20191107_Mix2_CE20.mzML", backend = MsBackendMzR())
sps

## How many spectra?
length(sps)
```

- `Spectra`: a list of MS spectra.
- *spectra variables*: values characterizing an MS spectrum
  - m/z and intensity values
  - MS level
  - retention time
    - precursor m/z
    - ...
- *Example*: access spectra variables:
```{r}
## Using accessor method
msLevel(sps)
    
## Easier: with $
sps$msLevel
    
## How many MS 1 and MS 2 spectra?
table(sps$msLevel)    
    
## m/z and intensity values
mz(sps)
intensity(sps)
    
## What variables do we have?
spectraVariables(sps)
```

- *Example*: add new spectra variables:
```{r}
sps$new_variable <- "A"

sps$new_variable
```
- Possibility to add arbitrary annotations to each spectrum.

- *Example*: subset a `Spectra` object:
```{r}
## get all MS level 2 spectra
ms2 <- filterMsLevel(sps, 2)
ms2
    
## or [
ms2 <- sps[sps$msLevel == 2]
ms2
```


## The concept of the `MsBackend`

- `MsBackend` classes: store and provide the spectrum data.
- *Example: `MsBackendMzR`: what data is actually stored?
```{r}
slotNames(sps)
    
## the backend *provides* the data
sps@backend
    
## what data is actually in memory?
sps@backend@spectraData
print(object.size(sps), units = "MB")
```

- *Example*: change backend to `MsBackendDataFrame`.
```{r}
sps <- setBackend(sps, MsBackendDataFrame())
    
## what's the size?
print(object.size(sps), unit = "MB")
```

- With `MsBackendDataFrame`, all data is in memory, also m/z and intensity
  values.
- Advantage of the backend concept?
  - Separation of data storage logic and spectrum data processing functionality.
  - Can have backends for differet data types or resources.
  - Can switch between backends depending on needs: low memory footprint, faster
    performance ...


## `MsBackendHmdb`: import data from HMDB

- Human metabolome database: MS2 spectra available as individual xml file.
- `MsBackendHmdbXml`: implemented in its own R package, reads data from xml.
- *Example*: load a subset of MS/MS spectra from HMDB
```{r}
library(MsBackendHmdb)
hmdb <- Spectra(dir("data/hmdb", full.names = TRUE),
                source = MsBackendHmdbXml())
hmdb

spectraVariables(hmdb)
head(hmdb$splash)
```


## Match experimental spectrum against HMDB

- *Goal*: match experimental MS2 spectrum against *reference* spectra from the
  Human Metabolome Database (HMDB).
- *Example*: identify MS2 spectra with precursor m/z that matches the [M+H]+ ion
  of Histidine:
```{r}
mz_hist <- 156.07675
ms2_hist <- sps %>%
    filterMsLevel(2) %>%
    filterPrecursorMz(mz = mz_hist + ppm(c(-mz_hist, mz_hist), 20))
ms2_hist

## it's a `Spectra`, even if there is only a single spectrum
plot(mz(ms2_hist)[[1]], intensity(ms2_hist)[[1]], type = "h",
     xlab = "m/z", ylab = "intensity")

## Is the data centroided?
centroided(ms2_hist) 
```

- *Example*: do some data manipulations: centroiding and removing low intensity
  peaks.
```{r}
ms2_hist <- ms2_hist %>%
    pickPeaks() %>%
    removePeaks(threshold = 500) %>%
    clean(all = TRUE)
ms2_hist

plot(mz(ms2_hist)[[1]], intensity(ms2_hist)[[1]], type = "h",
     xlab = "m/z", ylab = "intensity") 
```

- *Example*: match against reference database. 
```{r}
res <- compareSpectra(ms2_hist, hmdb, ppm = 40)

head(res)

## Select best matches
hmdb_hist <- hmdb[res > 0.7]
hmdb_hist

## plot experimenta against database spectrum
plot(3, 3, pch = NA, xlim = range(mz(hmdb_hist)[[1]], mz(ms2_hist)[[1]]),
     ylim = c(-1, 1), xlab = "m/z", ylab = "intensity")
abline(h = 0)
points(mz(hmdb_hist)[[1]], intensity(hmdb_hist)[[1]], type = "h")
points(mz(ms2_hist)[[1]], -intensity(ms2_hist)[[1]] /
                          max(intensity(ms2_hist)[[1]]),
       type = "h")
```


## The future

- Slowly replace `MSnbase` with the new packages.
- Slowly integrate `Spectra`, `Chromatograms` into `xcms`.
- Looking forward to contributions from the community.

## Opportunities

- We're open for contributions! See <https://www.rformassspectrometry.org> for
  more information.

### Implement own backends

- [ ] Backend to read/write *mgf* files (Michael Witting?).
- [ ] Backend to read/write MassBank files (Michael Witting?, Emma Schymanski?).


### Plotting functionality

- [ ] plotting functions for `Spectra` objects (base R graphics).
- [ ] ggplot-based plotting should ideally go into a separate package.


### Integrate new packages in your own workflow/software

I'd suggest you switch sooner than later.

